Power simulation section

### 2023-08-18 docs/manuscript.Rmd
add more references

wordcount         : "6264"

@@

<!-- Power simulations are hard! (especially for GLMMs) -->
Power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes needed to achieve sufficient statistical power for results with high evidencial value.
While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of (generalized) linear mixed models (GLMMs; also called multilevel models) in human-AI interaction research introduces greater complexity to power analysis.
For GLMMs, analytic options to derive sample sizes are not available for general research designs.
An increasing number of tutorials describe how to perform power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @murayamaSummarystatisticsbasedPowerAnalysis2022; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021].
However, most tutorials discuss only linear mixed models (LMMs) and focus on the most common designs, limiting the guidance available for researchers facing more complex study designs [@brownIntroductionLinearMixedEffects2021].
Moreover, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that meta-analytic results or data from pilot studies are available to determine plausible values for all model parameters.
In contrast, simulations are often required, necessitating assumptions for all parameters in the model [@maxwellSampleSizePlanning2008].
Therefore, running sample size calculations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing their significance in ensuring robust and meaningful findings.
By incorporating GLMMs and power analysis into research planning, researchers can enhance the replicability and credibility of their work, ultimately contributing to the advancement of science [@yarkoniGeneralizabilityCrisis2022].

@@

<!-- describe the lme4 package and how it works-->
In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015]. The lme4 package is a state-of-the-art tool for fitting frequentist GLMMs (for Bayesian GLMMs, the brms R package is currently the most prominent option [@burknerBrmsPackageBayesian2017]).
The lme4 package also includes a function called `simulate` that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling power simulations and other related analyses.
However, the model parameterization used by the lme4 package is quite technical, making it more difficult for applied researchers to determine whether their specified population model implies plausible associations in their simulated data.
Therefore in this tutorial, we simulate data for GLMMs from first principles to assist applied researchers in better understanding all model assumptions, and then use lme4 to analyze the simulated data sets.[^simr]

@@

<!-- GLMM intro -->
LMMs (Linear Mixed Models) and GLMMs (Generalized Linear Mixed Models) are powerful statistical frameworks that handle complex data structures by incorporating both fixed and random effects. 
LMMs are extensions of linear regression models that account for correlated data and hierarchical structures. They are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). LMMs allow for the modeling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects.
GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution, but instead belongs to a different distribution family, such as binomial, Poisson, or gamma. GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

<!-- Relevance of GLMMs -->
GLMMs are gaining increasing popularity in the field of Human-AI interaction research. As the complexity of studying human interactions with artificial intelligence systems grows, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs. GLMMs offer a flexible framework for analyzing data with non-normal and categorical outcomes, accounting for both fixed and random effects. This versatility makes GLMMs particularly suitable for investigating various aspects of Human-AI interaction, such as user preferences, trust, engagement, and performance. By incorporating GLMMs into their analyses, researchers in the field of Human-AI interaction can obtain more robust and comprehensive insights into the intricate dynamics between humans and AI systems, leading to a deeper understanding of the psychological and behavioral aspects involved.

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$.
$$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$
Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$.
$$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$
If the dependent variable $Y$ is a binary variable with values $0$ or $1$, the conditional expected value is equivalent to the probability:
$$
P_{si} = P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$
In our case study, $P_{si}$ is the conditional probability that subject $s$ gives the correct response to item $i$.

@@

Because all hypotheses tested simultaneously with the `glht` function must have the same direction, we flip the sign of inequalities three and four by multiplying them with $-1$.
The multcomp package automatically adjusts p-values when multiple hypotheses are tested simultaneously.
However, the combined null hypothesis in our exemplary research question should only be rejected if **all** individual null hypotheses are rejected.
In such cases, the error probabilities do not accumulate and we would waste power when correcting for multiple testing.
Thus, we request unadjusted p-values by setting `test = univariate()` in the `summary` command.
With a standard significance level of $\alpha = 0.05$, we would not reject all four null hypotheses (the p-value for hypothesis $H_{03}$ is not significant) and therefore also not reject the combined null hypothesis for this simulated dataset.
Note that this decision would be wrong, because we have simulated the data such that the combined alternative hypothesis is actually true in the population.

@@

The well known inequality of conditional and marginal effects in GLMMs makes their interpretation more difficult, however, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values:
When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

Much has been written on the optimal amount of power to target in empirical research.
The most prominent heuristic is to target a power of 0.8 (when combined with a type I error rate of $\alpha = 0.05$), but depending on the research goals of the study, there are often good reasons to move away from this standard.
When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other, based on practical considerations.
For the sake of the example, let the targeted power be indeed about 0.8 to detect an effect of the expected size implied by our data simulation.
This could be achieved by collecting data of 200 subjects (about 25% of which will be experts) each completing the same 50 items (with advice present in about 67% of cases which is correct in about 80% of cases with present advice).
If collecting 200 subjects is not practically feasible, it would also be possible to recruit 150 subjects but increase the length of the experiment to over 70 items.
However, 70 items might take to long to complete for the doctors participating in the study, who probably all have a busy schedule.
The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items, if it is feasible to recruit an even higher number of subjects (> 250, to be determined by additional rounds of power analysis).
Design parameters which also affect power and which could be investigated in the simulation to find a more optimal trade-off are the ratio of experts, the frequency of whether advice is presented and whether it is correct.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge.
Domain knowledge can be considered a vague theoretical model about the data-generating process that is less formal and can only be accessed by a back-and-forth exchange in which domain experts assess the plausibility of simulated data.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower and the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero.
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022].
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression on how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g., analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).

>>>>

wordcount         : "6551"

@@

<!-- Power simulations are hard! (especially for GLMMs) -->
Power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes needed to achieve sufficient statistical power for results with high evidencial value.
While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of (generalized) linear mixed models (GLMMs; also called multilevel models) in human-AI interaction research introduces greater complexity to power analysis.
For GLMMs, analytic options to derive sample sizes are not available for general research designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022].
An increasing number of tutorials describe how to perform power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021].
However, most tutorials discuss only linear mixed models (LMMs) and focus on the most common designs, limiting the guidance available for researchers facing more complex study designs [@brownIntroductionLinearMixedEffects2021].
Moreover, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that meta-analytic results or data from pilot studies are available to determine plausible values for all model parameters.
In contrast, simulations are often required, necessitating assumptions for all parameters in the model [@maxwellSampleSizePlanning2008].
Therefore, running sample size calculations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing their significance in ensuring robust and meaningful findings.
By incorporating GLMMs and power analysis into research planning, researchers can enhance the replicability and credibility of their work, ultimately contributing to the advancement of science [@yarkoniGeneralizabilityCrisis2022].

@@

<!-- describe the lme4 package and how it works-->
In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015]. The lme4 package is a state-of-the-art tool for fitting frequentist GLMMs.[^brms]

[^brms]: For Bayesian GLMMs, the brms R package is currently the most prominent option [@burknerBrmsPackageBayesian2017].

The lme4 package also includes a function called `simulate` that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling power simulations and other related analyses.
However, the model parameterization used by the lme4 package is quite technical, making it more difficult for applied researchers to determine whether their specified population model implies plausible associations in their simulated data.
Therefore in this tutorial, we simulate data for GLMMs from first principles to assist applied researchers in better understanding all model assumptions, and then use lme4 to analyze the simulated data sets.[^simr]

@@

<!-- GLMM intro -->
LMMs (Linear Mixed Models) and GLMMs (Generalized Linear Mixed Models) are powerful statistical frameworks that handle complex data structures by incorporating both fixed and random effects [@fahrmeirRegressionModelsMethods2021]. 
LMMs are extensions of linear regression models that account for correlated data and hierarchical structures. They are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). LMMs allow for the modeling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects.
GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution, but instead belongs to a different distribution family, such as binomial, Poisson, or gamma. GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

<!-- Relevance of GLMMs -->
GLMMs are gaining increasing popularity in the field of Human-AI interaction research. As the complexity of studying human interactions with artificial intelligence systems grows, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs offer a flexible framework for analyzing data with non-normal and categorical outcomes, accounting for both fixed and random effects. This versatility makes GLMMs particularly suitable for investigating various aspects of Human-AI interaction, such as user preferences, trust, engagement, and performance. By incorporating GLMMs into their analyses, researchers in the field of Human-AI interaction can obtain more robust and comprehensive insights into the intricate dynamics between humans and AI systems, leading to a deeper understanding of the psychological and behavioral aspects involved.

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021].
$$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$
Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$.
$$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$
If the dependent variable $Y$ is a binary variable with values $0$ or $1$, the conditional expected value is equivalent to the probability:
$$
P_{si} = P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$
In our case study, $P_{si}$ is the conditional probability that subject $s$ gives the correct response to item $i$.

@@

Because all hypotheses tested simultaneously with the `glht` function must have the same direction, we flip the sign of inequalities three and four by multiplying them with $-1$.
The multcomp package automatically adjusts p-values when multiple hypotheses are tested simultaneously [@hothornSimultaneousInferenceGeneral2008].
However, the combined null hypothesis in our exemplary research question should only be rejected if **all** individual null hypotheses are rejected [i.e. intersection-union setting\; @dmitrienkoTraditionalMultiplicityAdjustment2013].
In such cases, the error probabilities do not accumulate and we would waste power when correcting for multiple testing.
Thus, we request unadjusted p-values by setting `test = univariate()` in the `summary` command.
With a standard significance level of $\alpha = 0.05$, we would not reject all four null hypotheses (the p-value for hypothesis $H_{03}$ is not significant) and therefore also not reject the combined null hypothesis for this simulated dataset.
Note that this decision would be wrong, because we have simulated the data such that the combined alternative hypothesis is actually true in the population.

@@

The well known inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult, however, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values:
When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

Much has been written on the optimal amount of power to target in empirical research.
The most prominent heuristic is to target a power of 0.8 (when combined with a type I error rate of $\alpha = 0.05$), but depending on the research goals of the study, there are often good reasons to move away from this standard [@lakensJustifyYourAlpha2018; @lakensSampleSizeJustification2022].
When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other, based on practical considerations.
For the sake of the example, let the targeted power be indeed about 0.8 to detect an effect of the expected size implied by our data simulation.
This could be achieved by collecting data of 200 subjects (about 25% of which will be experts) each completing the same 50 items (with advice present in about 67% of cases which is correct in about 80% of cases with present advice).
If collecting 200 subjects is not practically feasible, it would also be possible to recruit 150 subjects but increase the length of the experiment to over 70 items.
However, 70 items might take to long to complete for the doctors participating in the study, who probably all have a busy schedule.
The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items, if it is feasible to recruit an even higher number of subjects (> 250, to be determined by additional rounds of power analysis).
Design parameters which also affect power and which could be investigated in the simulation to find a more optimal trade-off are the ratio of experts, the frequency of whether advice is presented and whether it is correct.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge.
Domain knowledge can be considered a vague theoretical model about the data-generating process that is less formal and can only be accessed by a back-and-forth exchange in which domain experts assess the plausibility of simulated data.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower and the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero [@kingPointMinimalImportant2011].
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022].
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression on how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g., analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).
